---
header-includes: |
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhead[LO,L]{\textbf{MTH-321 AB} | Spring 2026}
  \fancyhead[RO,R]{\textbf{Homework Set 4} | Pg. \thepage}
  \fancyfoot[CO,C]{}
  \usepackage{float}
  \usepackage{multirow}
  \usepackage{enumitem}
  \usepackage{tasks}
output:
  bookdown::pdf_document2:
    toc: no
    number_section: no
urlcolor: blue
link-citations: yes
bibliography: ../../../references.bib
csl: ../../../apa.csl
geometry: "left=0.50in,right=0.50in,top=0.75in,bottom=0.75in"
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source(file.path("../../R/functions-mth321.R"))
```

**Name:**

\noindent\rule{\textwidth}{1pt}
**Instructions:** Homework is graded primarily on correctness and completion.

* Write your solutions on a separate sheet of paper.
* Include all key steps and explanations that justify your answers, unless it says otherwise.
* If you print this homework set, you may write directly on the page where appropriate.
* Attach all supporting files (e.g. \texttt{.py} or \texttt{.ipynb} files) when needed.

\noindent\rule{\textwidth}{1pt}

## Homework Set 4

### 1.

Let $x$ and $y$ be the dependent variables with $t$ as the independent variable. For each system use the given information to:

  (i) determine the general and specific solution in vector form, 
  (ii) identify the correct direction field, shown below, by matching the eigenvectors,
  (iii) draw the specific solution on the direction field, and 
  (iv) use a graphing technology to plot the specific solution with respect to the $t$ axis.
  
Then, describe the similarities and differences between how solution behavior appears on the phase plane and how it appears when graphed as a function of time.

\begin{tasks}
  \task $\displaystyle \begin{bmatrix} x' \\ y' \end{bmatrix} = \begin{bmatrix} 3 & 4 \\ -1 & 7 \end{bmatrix} \begin{bmatrix} x \\ y \end{bmatrix}, \hspace{10px} \vec{y}(0) = \begin{bmatrix} 1 \\ 1 \end{bmatrix}; \hspace{10px} \lambda_1 = 5, \hspace{10px} \vec{v_1} = \begin{bmatrix} 2 \\ 1 \end{bmatrix}$
  \vspace{10px}
  \task $\displaystyle \begin{bmatrix} x' \\ y' \end{bmatrix} = \begin{bmatrix} 1 & 2 \\ -4 & 5 \end{bmatrix} \begin{bmatrix} x' \\ y' \end{bmatrix}, \hspace{10px} \vec{y}(0) = \begin{bmatrix} -1 \\ 1 \end{bmatrix}; \hspace{10px} \begin{aligned} \lambda_1 = & 3+2i \\ \lambda_2 = & 3-2i, \end{aligned} \hspace{10px} \vec{v_1} = \begin{bmatrix} 1 \\ 2 \end{bmatrix} + i \begin{bmatrix} -1 \\ 0 \end{bmatrix}, \hspace{10px} \vec{v_2} = \begin{bmatrix} 1 \\ 2 \end{bmatrix} + i \begin{bmatrix} 1 \\ 0 \end{bmatrix}$
  \vspace{10px}
\end{tasks}

\begin{minipage}[t!]{0.49\textwidth}
```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width="100%", fig.width=3,fig.height=3}
ode <- function(x,y,parameters){
  with(as.list(c(parameters)),{
    dx <- 1*x + 2*y
    dy <- -4*x + 5*y
    return(list(dx,dy))
  })
}

PhasePortrait(ode,params=c(),
              xlim=c(-2,2),ylim=c(-2,2),res=15,
              scale=0.25,rad=0.25,unit=TRUE,
              col="black",xlab="x",ylab="y",title="")  +
  geom_point(data=tibble(x=0,y=0),aes(x=x,y=y),color="red") + 
  ggtitle("A")
```
\end{minipage}
\begin{minipage}[t!]{0.49\textwidth}
```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width="100%", fig.width=3,fig.height=3}
ode <- function(x,y,parameters){
  with(as.list(c(parameters)),{
    dx <- 3*x + 4*y
    dy <- -1*x + 7*y
    return(list(dx,dy))
  })
}

PhasePortrait(ode,params=c(),
              xlim=c(-2,2),ylim=c(-2,2),res=15,
              scale=0.25,rad=0.25,unit=TRUE,
              col="black",xlab="x",ylab="y",title="")  +
  geom_point(data=tibble(x=0,y=0),aes(x=x,y=y),color="red") + 
  ggtitle("B")
```
\end{minipage}

### 2.

Consider the following systems of ODEs with their corresponding direction fields:

\begin{minipage}[t!]{0.49\textwidth}
\begin{center}\underline{System A}\end{center}
\vspace{10px}
$$
\begin{aligned}
x' & = 2x + y \\
y' & = x + 2y
\end{aligned}
$$
```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width="100%", fig.width=3,fig.height=3}
ode <- function(x,y,parameters){
  with(as.list(c(parameters)),{
    dx <- 2*x + y
    dy <- x + 2*y
    return(list(dx,dy))
  })
}

PhasePortrait(ode,params=c(),
              xlim=c(-2,2),ylim=c(-2,2),res=17,
              scale=0.20,rad=0.225,unit=TRUE,
              col="black",xlab="x",ylab="y",title="")  +
  geom_point(data=tibble(x=0,y=0),aes(x=x,y=y),color="red")
```
\end{minipage}
\begin{minipage}[t!]{0.49\textwidth}
\begin{center}\underline{System B}\end{center}
\vspace{10px}
$$
\begin{aligned}
x' & = 2x - \frac{1}{3} xy \\
y' & = 3y - \frac{1}{2} xy
\end{aligned}
$$
```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width="100%", fig.width=3,fig.height=3}
ode <- function(x,y,parameters){
  with(as.list(c(parameters)),{
    dx <- 2*x - (1/3)*x*y
    dy <- 3*y - (1/2)*x*y
    return(list(dx,dy))
  })
}

PhasePortrait(ode,params=c(),
              xlim=c(-3,9),ylim=c(-3,9),res=17,
              scale=0.50,rad=0.1,unit=TRUE,
              col="black",xlab="x",ylab="y",title="")  +
  geom_point(data=tibble(x=0,y=0),aes(x=x,y=y),color="red") + 
  geom_point(data=tibble(x=6,y=6),aes(x=x,y=y),color="red")
```
\end{minipage}

\begin{tasks}
  \task Determine the Jacobian matrices $J_A(x,y)$ and $J_B(x,y)$ for systems $A$ and $B$ respectively. Compare the matrices and identify which system is linear and non-linear. Then, explain the key indicator of the system's linearity based on the matrices.
  \task For each system, determine the equilibriums and evaluate their Jacobian matrix at those points. Then, explain the purpose of the Jacobian matrix in the process of linearization.
  \task For each system, determine the eigenvalues of each Jacobian matrix. Then, explain what the eigenvalues describe regarding the behavior of solutions around each equilibrium.
  \task For the linear system only:
    \begin{itemize}
      \item[i.] Determine the "straight-line" solutions using the isocline $\displaystyle \frac{y'}{x'} = \frac{y}{x}$ and compare these solutions to the eigenvalues for the linear system computed in Part (d). Describe your observations.
      \item[ii.] Draw these "straight-line" solutions on the system's corresponding direction field. Describe the behavior of solutions around the equilibrium.
      \item[iii.] Are the properties of the eigenvalues consistent with your observations at the equilibrium?
    \end{itemize}
  \task For the non-linear system only:
    \begin{itemize}
      \item[i.] Explain why the linearization process is necessary to determine the behavior of solutions around an equilibrium.
      \item[ii.] Draw at least four specific solution curves on the system's corresponding direction field. Describe the behavior of solutions around each equilibrium.
      \item[iii.] Are the properties of the eigenvalues consistent with your observations at each equilibrium?
    \end{itemize}
\end{tasks}

\newpage

### 3.

Let $x$ and $y$ be the dependent variables with $t$ as the independent variable. For each of the following homogeneous linear system of 1st-order ODEs: 
  
  (i) determine the general solution, 
  (ii) on its corresponding direction field, identify the "straight-line" solutions if any, 
  (iii) on the same direction field, draw at least four specific solution curves, and 
  (iv) identify the equilibrium class (including the relevant subclass) and its stability.

\begin{tasks}(2)
  \task
    $\displaystyle \begin{aligned} x' = & -2x-1y \\ y' = & -2y \end{aligned}$
```{r echo=FALSE, fig.align='left', message=FALSE, warning=FALSE, out.width="100%", fig.width=3,fig.height=3}
ode <- function(x,y,parameters){
  with(as.list(c(parameters)),{
    dx <- -2*x - 1*y
    dy <- 0*x - 2*y
    return(list(dx,dy))
  })
}

PhasePortrait(ode,params=c(),
              xlim=c(-2,2),ylim=c(-2,2),res=15,
              scale=0.25,rad=0.25,unit=TRUE,
              col="black",xlab="x",ylab="y",title="")  +
  geom_point(data=tibble(x=0,y=0),aes(x=x,y=y),color="red")
```
  \task 
    $\displaystyle \begin{aligned} x' = & -4x-2y \\ y' = & -x-3y \end{aligned}$
```{r echo=FALSE, fig.align='left', message=FALSE, warning=FALSE, out.width="100%", fig.width=3,fig.height=3}
ode <- function(x,y,parameters){
  with(as.list(c(parameters)),{
    dx <- -4*x - 2*y
    dy <- -1*x - 3*y
    return(list(dx,dy))
  })
}

PhasePortrait(ode,params=c(),
              xlim=c(-2,2),ylim=c(-2,2),res=15,
              scale=0.25,rad=0.25,unit=TRUE,
              col="black",xlab="x",ylab="y",title="")  +
  geom_point(data=tibble(x=0,y=0),aes(x=x,y=y),color="red")
```
  \task
    $\displaystyle \begin{aligned} x' = & 4y \\ y' = & x \end{aligned}$
```{r echo=FALSE, fig.align='left', message=FALSE, warning=FALSE, out.width="100%", fig.width=3,fig.height=3}
ode <- function(x,y,parameters){
  with(as.list(c(parameters)),{
    dx <- 0*x + 4*y
    dy <- 1*x + 0*y
    return(list(dx,dy))
  })
}

PhasePortrait(ode,params=c(),
              xlim=c(-2,2),ylim=c(-2,2),res=15,
              scale=0.25,rad=0.25,unit=TRUE,
              col="black",xlab="x",ylab="y",title="")  +
  geom_point(data=tibble(x=0,y=0),aes(x=x,y=y),color="red")
```
  \task 
    $\displaystyle \begin{bmatrix} x' \\ y' \end{bmatrix} = \begin{bmatrix} 4 & -2 \\ -2 & 1 \end{bmatrix} \begin{bmatrix} x \\ y \end{bmatrix}$
```{r echo=FALSE, fig.align='left', message=FALSE, warning=FALSE, out.width="100%", fig.width=3,fig.height=3}
ode <- function(x,y,parameters){
  with(as.list(c(parameters)),{
    dx <- 4*x - 2*y
    dy <- -2*x + 1*y
    return(list(dx,dy))
  })
}

PhasePortrait(ode,params=c(),
              xlim=c(-2,2),ylim=c(-2,2),res=14,
              scale=0.25,rad=0.25,unit=TRUE,
              col="black",xlab="x",ylab="y",title="") +
  geom_abline(slope=2,intercept=0,color="red",size=1)
```
  \task
    $\displaystyle \begin{bmatrix} x' \\ y' \end{bmatrix} = \begin{bmatrix} -1 & -1 \\ 2 & 1 \end{bmatrix} \begin{bmatrix} x \\ y \end{bmatrix}$
```{r echo=FALSE, fig.align='left', message=FALSE, warning=FALSE, out.width="100%", fig.width=3,fig.height=3}
ode <- function(x,y,parameters){
  with(as.list(c(parameters)),{
    dx <- -1*x - 1*y
    dy <- 2*x + 1*y
    return(list(dx,dy))
  })
}

PhasePortrait(ode,params=c(),
              xlim=c(-2,2),ylim=c(-2,2),res=15,
              scale=0.25,rad=0.25,unit=TRUE,
              col="black",xlab="x",ylab="y",title="")  +
  geom_point(data=tibble(x=0,y=0),aes(x=x,y=y),color="red")
```
  \task
    $\displaystyle \begin{bmatrix} x' \\ y' \end{bmatrix} = \begin{bmatrix} -1 & 2 \\ -2 & -1 \end{bmatrix} \begin{bmatrix} x \\ y \end{bmatrix}$
```{r echo=FALSE, fig.align='left', message=FALSE, warning=FALSE, out.width="100%", fig.width=3,fig.height=3}
ode <- function(x,y,parameters){
  with(as.list(c(parameters)),{
    dx <- -1*x + 2*y
    dy <- -2*x - 1*y
    return(list(dx,dy))
  })
}

PhasePortrait(ode,params=c(),
              xlim=c(-2,2),ylim=c(-2,2),res=15,
              scale=0.25,rad=0.25,unit=TRUE,
              col="black",xlab="x",ylab="y",title="") +
  geom_point(data=tibble(x=0,y=0),aes(x=x,y=y),color="red")
```
  \task 
    $\displaystyle \begin{bmatrix} x' \\ y' \end{bmatrix} = \begin{bmatrix} 2 & -3 \\ 3 & 2 \end{bmatrix} \begin{bmatrix} x \\ y \end{bmatrix}$
```{r echo=FALSE, fig.align='left', message=FALSE, warning=FALSE, out.width="100%", fig.width=3,fig.height=3}
ode <- function(x,y,parameters){
  with(as.list(c(parameters)),{
    dx <- 2*x - 3*y
    dy <- 3*x + 2*y
    return(list(dx,dy))
  })
}

PhasePortrait(ode,params=c(),
              xlim=c(-2,2),ylim=c(-2,2),res=15,
              scale=0.25,rad=0.25,unit=TRUE,
              col="black",xlab="x",ylab="y",title="") +
  geom_point(data=tibble(x=0,y=0),aes(x=x,y=y),color="red")
```
  \task 
    $\displaystyle \begin{aligned} x' = & -2x \\ y' = & -2y \end{aligned}$
```{r echo=FALSE, fig.align='left', message=FALSE, warning=FALSE, out.width="100%", fig.width=3,fig.height=3}
ode <- function(x,y,parameters){
  with(as.list(c(parameters)),{
    dx <- -2*x + 0*y
    dy <- 0*x + -2*y
    return(list(dx,dy))
  })
}

PhasePortrait(ode,params=c(),
              xlim=c(-2,2),ylim=c(-2,2),res=15,
              scale=0.25,rad=0.25,unit=TRUE,
              col="black",xlab="x",ylab="y",title="") +
  geom_point(data=tibble(x=0,y=0),aes(x=x,y=y),color="red")
```
\end{tasks}

\newpage

### 4. Challenge Problem

Consider the following homogeneous linear system of 1st-order ODEs.
$$
\begin{aligned}
  x' = & - \alpha x + \beta y\\
  y' = & \beta x -\alpha y
\end{aligned}
$$

where $\alpha > 0$ and $\beta > 0$.

\begin{tasks}
  \task Determine the equation for the $x$-nullcline of the system of ODEs with terms of $\alpha$ and $\beta$.
  \task Determine the equation for the $y$-nullcline of the system of ODEs with terms of $\alpha$ and $\beta$.
  \task Determine the "straight-line" solutions (eigenvalues and eigenvectors) with terms of $\alpha$ and $\beta$ in two ways:
    \begin{itemize}
      \item[i.] Using the isocline $\displaystyle \frac{y'}{x'} = \frac{y}{x}$.
      \item[ii.] Using the matrix-vector form $\displaystyle \begin{bmatrix} x' \\ y' \end{bmatrix} = \begin{bmatrix} -\alpha & \beta \\ \beta & -\alpha \end{bmatrix} \begin{bmatrix} x \\ y \end{bmatrix}$.
    \end{itemize}
    Then, compare the results obtained by each method. Discuss any similarities or differences.
  \task Create complete and well represented phase portraits by hand with at least four specific solution curves for cases when $\alpha = \beta$, $\alpha < \beta$, and $\alpha > \beta$.
  \task Describe the behavior of solutions around the equilibrium for cases when $\alpha = \beta$, $\alpha < \beta$, and $\alpha > \beta$.
\end{tasks}
