---
header-includes: |
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhead[LO,L]{\textbf{MTH-321 AB} | Spring 2026}
  \fancyhead[RO,R]{\textbf{Worksheet 11} | Pg. \thepage}
  \fancyfoot[CO,C]{}
  \usepackage{float}
  \usepackage{multirow}
output:
  bookdown::pdf_document2:
    toc: no
    number_section: no
urlcolor: blue
link-citations: yes
bibliography: ../../../references.bib
csl: ../../../mbe.csl
geometry: "left=0.50in,right=0.50in,top=0.75in,bottom=0.75in"
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source(file.path("../../R/functions-mth321.R"))
```

**Name:**

**Collaborators:**

\noindent\rule{\textwidth}{1pt}
**Instructions:** Worksheets are graded mostly on completion and partially on correctness. Please write complete solutions showing explanations and key steps to the following problems, unless it says otherwise.
\noindent\rule{\textwidth}{1pt}

## Modeling Solution Mixing

### 1. Mixing Problem

The amount of a substance $C$ in a solution (a mixture of a fluid and a substance) at time $t$ is modeled by the ODE $$C' = 6 - \frac{2C}{10 + Mt},$$ where $C'$ is the substance rate of change, the constant term $6$ is the rate of incoming solution, the term $- \frac{2C}{100 + t}$ represents the rate at which the substance flows out, with the numerator corresponding to the amount of substance and the denominator representing the fluid volume at time $t$ that changes at a rate of $M$.

a. Use the blank axis below to draw complete and well-represented slope fields as the parameter $M$ changes for the cases:

\begin{minipage}[t][0.32\textheight]{0.49\textwidth}
  \begin{itemize}
    \item[i.] $M = -1$
  \end{itemize}
```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width="100%", fig.width=4, fig.height=3}
# Create an empty data frame
empty_df <- data.frame(x = numeric(0), y = numeric(0))

# Generate the empty plot with grid lines
ggplot(empty_df, aes(x = x, y = y)) +
  geom_blank() +
  geom_vline(xintercept=0, color="black") +
  geom_hline(yintercept=0, color="black") +
  xlim(0,20) +
  xlab("t") +
  ylim(0,60) +
  ylab("C") +
  theme_minimal()
```
\end{minipage}
\hfill\vline\hfill
\begin{minipage}[t][0.32\textheight]{0.49\textwidth}
  \begin{itemize}
    \item[ii.] $M = 0$
  \end{itemize}
```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width="100%", fig.width=4, fig.height=3}
# Create an empty data frame
empty_df <- data.frame(x = numeric(0), y = numeric(0))

# Generate the empty plot with grid lines
ggplot(empty_df, aes(x = x, y = y)) +
  geom_blank() +
  geom_vline(xintercept=0, color="black") +
  geom_hline(yintercept=0, color="black") +
  xlim(0,20) +
  xlab("t") +
  ylim(0,60) +
  ylab("C") +
  theme_minimal()
```
\end{minipage}

b. For each $M$ case in Part (a), describe the general behavior of the substance using the slope fields. Assume that the initial substance starts at $C(0) > 0$.
\vfill

\newpage

### 2. Solving a Mixing Problem

Continued from Problem (1). Consider the initial value problem, $$C' = 6 - \frac{2C}{10 + t}, \quad C(0) = 50.$$

\begin{minipage}[t][\textheight-\pagetotal]{0.49\textwidth}
  \begin{itemize}
    \item[a.] Use the method of integrating factors to determine the general solution.
    \vfill
    \item[b.] Determine the specific solution.
    \vfill
  \end{itemize}
\end{minipage}
\hfill\vline\hfill
\begin{minipage}[t][\textheight-\pagetotal]{0.49\textwidth}
  \begin{itemize}
    \item[c.] Use the blank axis below to draw a complete and well represented slope field of the ODE. Then, draw the specific solution on the same field.
```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width="100%", fig.width=4, fig.height=3}
# Create an empty data frame
empty_df <- data.frame(x = numeric(0), y = numeric(0))

# Generate the empty plot with grid lines
ggplot(empty_df, aes(x = x, y = y)) +
  geom_blank() +
  geom_vline(xintercept=0, color="black") +
  geom_hline(yintercept=0, color="black") +
  xlim(0,20) +
  xlab("t") +
  ylim(0,60) +
  ylab("C") +
  theme_minimal()
```
    \item[d.] Describe what happens when the inflow stops. That is when the the rate of incoming solution becomes $0$.
    \vfill
    \item[e.] Describe what happens when the tank overflows. At what point in time does the overflow happen if the tank capacity is $100$?
    \vfill
  \end{itemize}
\end{minipage}

<!-- \newpage

## **References**

::: {#refs}
:::

<br> -->
