---
header-includes: |
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhead[LO,L]{\textbf{MTH-321 AB} | Spring 2026}
  \fancyhead[RO,R]{\textbf{Worksheet 4} | Pg. \thepage}
  \fancyfoot[CO,C]{}
  \usepackage{float}
  \usepackage{multirow}
output:
  bookdown::pdf_document2:
    toc: no
    number_section: no
urlcolor: blue
link-citations: yes
bibliography: ../../../references.bib
csl: ../../../mbe.csl
geometry: "left=0.25in,right=0.25in,top=0.75in,bottom=0.75in"
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source(file.path("../../R/functions-mth321.R"))
```

**Name:**

**Collaborators:**

\noindent\rule{\textwidth}{1pt}
**Instructions:** Worksheets are graded mostly on completion and partially on correctness. Please write complete solutions showing explanations and key steps to the following problems, unless it says otherwise.
\noindent\rule{\textwidth}{1pt}

## Determine Equilibrium Stability

### 1. Equilibriums and Their Stability

In a 1st-order autonomous ODE $\frac{dy}{dt} = f(y)$, an equilibrium occurs at a point $y^*$ where $f(y^*) = 0$. The *stability* of the equilibrium $y^*$ depends on how nearby solutions behave:

* If solutions starting close to $y^*$ move toward it in time, the equilibrium is *stable*.
* If solutions starting close to $y^*$ move away from it in time, the equilibrium is *unstable*.

Consider the ODE, $\displaystyle \frac{dy}{dt} = y - y^2$.

\begin{minipage}[t][0.55\textheight]{0.49\textwidth}
\begin{itemize}
  \item[a.] Define $f(y)$. Then, determine all equilibrium solutions, $y(t) = y^*$.
  \vfill
  \item[c.] Use the blank axis below to complete the Phase Diagram of the given ODE and mark each equilibrium whether they are stable or unstable.
  \vspace{5px}
```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width="100%", fig.width=4,fig.height=3}
ode_logistic <- function(t,y,parameters){
  with(as.list(c(parameters)),{
    dy <- y - y^2
    return(list(dy))
  })
}

y_vec <- seq(-0.25,1.25,length.out=30)
dydt_vec <- ode_logistic(0,y_vec,NA)
ode_logistic_df <- data.frame(x = y_vec, y = dydt_vec[[1]])

ggplot(ode_logistic_df, aes(x = x, y = y)) +
  #geom_line(linewidth=1) +
  geom_blank() +
  geom_vline(xintercept=0, color="black") +
  geom_hline(yintercept=0, color="black") +
  xlab("y") +
  xlim(-0.25,1.25) +
  ylab("f") +
  ylim(-0.3,0.3) +
  theme_minimal()
```
\end{itemize}
\end{minipage}
\hfill\vline\hfill
\begin{minipage}[t][0.55\textheight]{0.49\textwidth}
\begin{itemize}
  \item[b.] Compute $\displaystyle \frac{\partial f}{\partial y}$ evaluated for all $y^*$. Then, mark each if positive or negative.
  \vfill
  \item[d.] What do you observe about the stability of the equilibriums and the signs of the partial derivatives computed in Part (b)?
  \vfill
\end{itemize}
\end{minipage}

\newpage

### 2. Semi-Stable Equilibriums

If solutions approach $y^*$ from one side but move away from it on the other side, the equilibrium is *semi-stable*.

Consider the ODE, $\displaystyle \frac{dy}{dt} = (y-2)(y+1)y^2$.

\begin{minipage}[t][0.88\textheight]{0.49\textwidth}
\begin{itemize}
  \item[a.] Define $f(y)$. Then, determine all equilibrium solutions, $y(t) = y^*$.
  \vfill
  \item[c.] Complete the Phase Diagram of the given ODE shown below and mark each equilibrium whether they are stable or unstable.
  \vspace{5px}
```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width="100%", fig.width=4,fig.height=3}
ode <- function(t,y,parameters){
  with(as.list(c(parameters)),{
    dy <- (y-2)*(y+1)*y^2
    return(list(dy))
  })
}

y_vec <- seq(-1.25,2.25,length.out=30)
dydt_vec <- ode(0,y_vec,NA)
ode_df <- data.frame(x = y_vec, y = dydt_vec[[1]])

ggplot(ode_df, aes(x = x, y = y)) +
  geom_line(linewidth=1) +
  geom_blank() +
  geom_vline(xintercept=0, color="black") +
  geom_hline(yintercept=0, color="black") +
  xlab("y") +
  ylab("f") +
  theme_minimal()
```
  \item[e.] Use the phase diagram to describe the behavior of solutions --including where it approaches-- given the following conditions. Explain your reasoning.
    \begin{itemize}
      \item[i.] $-1 < y(0) < 0$
      \vfill
      \item[ii.] $0 < y(0) < 2$
      \vfill
    \end{itemize}
\end{itemize}
\end{minipage}
\hfill\vline\hfill
\begin{minipage}[t][0.88\textheight]{0.49\textwidth}
\begin{itemize}
  \item[b.] Compute $\displaystyle \frac{\partial f}{\partial y}$ evaluated for all $y^*$. Then, mark each if positive or negative.
  \vfill
  \item[d.] What do you observe about the stability of the equilibriums and the signs of the partial derivatives computed in Part (b)?
  \vfill
  \item[f.] Show using $\displaystyle \frac{\partial^2 f}{\partial y^2}$ evaluated at a semi-stable equilibrium that the solutions around this equilibrium is either decreasing or increasing.
  \vfill
\end{itemize}
\end{minipage}

<!-- \newpage

## **References**

::: {#refs}
:::

<br> -->
