---
header-includes: |
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhead[LO,L]{\textbf{MTH-321 AB} | Spring 2026}
  \fancyhead[RO,R]{\textbf{Worksheet 10} | Pg. \thepage}
  \fancyfoot[CO,C]{}
  \usepackage{float}
  \usepackage{multirow}
output:
  bookdown::pdf_document2:
    toc: no
    number_section: no
urlcolor: blue
link-citations: yes
bibliography: ../../../references.bib
csl: ../../../mbe.csl
geometry: "left=0.25in,right=0.25in,top=0.75in,bottom=0.75in"
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source(file.path("../../R/functions-mth321.R"))
```

**Name:**

**Collaborators:**

\noindent\rule{\textwidth}{1pt}
**Instructions:** Worksheets are graded mostly on completion and partially on correctness. Please write complete solutions showing explanations and key steps to the following problems, unless it says otherwise.
\noindent\rule{\textwidth}{1pt}

## Modeling Population Change

### 1. Population Dynamics

The population proportion $P$ at time $t$, relative to the carrying capacity in an environment with limited resources, is modeled by the logistic growth/decay ODE, $$P' = P \left( 1 - P \right) + \beta,$$ where $P'$ is the rate of change of the population, $P$ is the natural rate of growth/decay, $\left( 1 - P \right)$ represents the rate of decay due to environmental pressure, and $\beta$ is the external growth/decay rate.

a. Use the blank axis below to draw complete and well-represented phase diagrams of the ODE as the parameter $\beta$ changes for cases:

\begin{minipage}[t][0.28\textheight]{0.30\textwidth}
  \begin{itemize}
    \item[i.] $\beta < -\frac{1}{4}$
  \end{itemize}
```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width="100%", fig.width=2.5, fig.height=2.5}
# Create an empty data frame
empty_df <- data.frame(x = numeric(0), y = numeric(0))

# Generate the empty plot with grid lines
ggplot(empty_df, aes(x = x, y = y)) +
  geom_blank() +
  geom_vline(xintercept=0, color="black") +
  geom_hline(yintercept=0, color="black") +
  xlim(0,1.5) +
  xlab("P") +
  ylim(-1.25,1.25) +
  ylab("P'") +
  theme_minimal()
```
\end{minipage}
\hfill\vline\hfill
\begin{minipage}[t][0.28\textheight]{0.30\textwidth}
  \begin{itemize}
    \item[ii.] $\beta = -\frac{1}{4}$
  \end{itemize}
```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width="100%", fig.width=2.5, fig.height=2.5}
# Create an empty data frame
empty_df <- data.frame(x = numeric(0), y = numeric(0))

# Generate the empty plot with grid lines
ggplot(empty_df, aes(x = x, y = y)) +
  geom_blank() +
  geom_vline(xintercept=0, color="black") +
  geom_hline(yintercept=0, color="black") +
  xlim(0,1.5) +
  xlab("P") +
  ylim(-1.25,1.25) +
  ylab("P'") +
  theme_minimal()
```
\end{minipage}
\hfill\vline\hfill
\begin{minipage}[t][0.28\textheight]{0.30\textwidth}
  \begin{itemize}
    \item[iii.] $\beta > -\frac{1}{4}$
  \end{itemize}
```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width="100%", fig.width=2.5, fig.height=2.5}
# Create an empty data frame
empty_df <- data.frame(x = numeric(0), y = numeric(0))

# Generate the empty plot with grid lines
ggplot(empty_df, aes(x = x, y = y)) +
  geom_blank() +
  geom_vline(xintercept=0, color="black") +
  geom_hline(yintercept=0, color="black") +
  xlim(0,1.5) +
  xlab("P") +
  ylim(-1.25,1.25) +
  ylab("P'") +
  theme_minimal()
```
\end{minipage}

b. For each $\beta$ case in Part (a), describe the general behavior of the population using the phase diagrams. Assume that the initial population starts at $0 < P(0) < 1$.
\vfill

\newpage

### 2. Logistic Growth Exact Solution

Continued from Problem (1). Consider the initial value problem, $$P' = P \left( 1 - P \right), \quad P(0) = \frac{1}{4}.$$

\begin{minipage}[t][\textheight-\pagetotal]{0.49\textwidth}
  \begin{itemize}
    \item[a.] Use the method of separation of variables to determine the general solution.
    \vfill
    \item[b.] Determine the specific solution.
    \vfill
  \end{itemize}
\end{minipage}
\hfill\vline\hfill
\begin{minipage}[t][\textheight-\pagetotal]{0.49\textwidth}
  \begin{itemize}
    \item[c.] Use the blank axis below to draw a complete and well represented slope field of the ODE. Then, draw the specific solution on the same field.
```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width="100%", fig.width=4, fig.height=3}
# Create an empty data frame
empty_df <- data.frame(x = numeric(0), y = numeric(0))

# Generate the empty plot with grid lines
ggplot(empty_df, aes(x = x, y = y)) +
  geom_blank() +
  geom_vline(xintercept=0, color="black") +
  geom_hline(yintercept=0, color="black") +
  xlim(0,2.50) +
  xlab("t") +
  ylim(-.25,1.25) +
  ylab("y") +
  theme_minimal()
```
    \item[d.] Determine the time it takes for the population to triple from its initial size.
\vfill
  \end{itemize}
\end{minipage}

<!-- \newpage

## **References**

::: {#refs}
:::

<br> -->
