---
header-includes: |
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhead[LO,L]{\textbf{MTH-321 AB} | Spring 2026}
  \fancyhead[RO,R]{\textbf{Worksheet 22} | Pg. \thepage}
  \fancyfoot[CO,C]{}
  \usepackage{float}
  \usepackage{multirow}
output:
  bookdown::pdf_document2:
    toc: no
    number_section: no
urlcolor: blue
link-citations: yes
bibliography: ../../../references.bib
csl: ../../../mbe.csl
geometry: "left=0.25in,right=0.25in,top=0.75in,bottom=0.75in"
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source(file.path("../../R/functions-mth321.R"))
```

**Name:**

**Collaborators:**

\noindent\rule{\textwidth}{1pt}
**Instructions:** Worksheets are graded mostly on completion and partially on correctness. Please write complete solutions showing explanations and key steps to the following problems, unless it says otherwise.
\noindent\rule{\textwidth}{1pt}

## Use Eigenvalues to Analyze Equilibrium Stability

### 1. Linearizations of 1st-Order Non-Linear System of ODEs

*Linearization* is the process of approximating the solution behaviors nearby and around an equilibrium.

Consider this *non-linear homogeneous system of 1st-order ODEs*:
$$
\begin{aligned}
x' & = 3x-\frac{7}{5}xy \\
y' & = -y+\frac{4}{5}xy
\end{aligned}
$$
with equilibrium solutions $(0,0)$ and $\left(\frac{5}{4},\frac{15}{7}\right)$.

a. The following tasks determines the Jacobian matrix $J(x,y)$, and defines $J_1$ and $J_2$ which is $J(x,y)$ evaluated at the equilibrium solutions. Determining the Jacobian Matrix is an important step for linearizing a non-linear system of 1st-order ODEs.

| Task | Show work here |
|---------|---------|
| Determine the partial derivatives $\displaystyle \frac{\partial f_1}{\partial x}$ and $\displaystyle \frac{\partial f_1}{\partial y}$ where $\displaystyle f_1 = 3x-\frac{7}{5}xy$. | |
|---------|---------|
| Determine the partial derivatives $\displaystyle \frac{\partial f_2}{\partial x}$ and $\displaystyle \frac{\partial f_2}{\partial y}$ where $\displaystyle f_2 = -y+\frac{4}{5}xy$. | |
|---------|---------|
| Construct the Jacobian matrix $\displaystyle J(x,y) = \begin{bmatrix} \frac{\partial f_1}{\partial x} & \frac{\partial f_1}{\partial y} \\ \frac{\partial f_2}{\partial x} & \frac{\partial f_2}{\partial y} \end{bmatrix}$. | |
|---------|---------|
| Determine $\displaystyle J_1 = J(0,0)$. | |
|---------|---------|
| Determine $\displaystyle J_2 = J\left(\frac{5}{4},\frac{15}{7}\right)$. | |

The *Jacobian Matrix* represents a tangent linear map that describes the local "slope" and directional behavior of an equilibrium solution.

\newpage

### 2. Eigenvalues for Determining Equilibrium Stability

Consider the same non-linear system given in Problem (1) and the Jacobian matrices $J_1$ and $J_2$.

The *eigenvalues* of the Jacobian matrix at an equilibrium can help us predict its stability because it represents the slope of the "straight-line" localized solutions around the equilibrium.

a. The following tasks determines the eigenvalues of $J_1$ and uses them to determine the stability of the equilibrium $(0,0)$.

| Task | Show work here |
|------|---------|
| Set $\displaystyle \text{det}\left(J_1 - \begin{bmatrix} \lambda & 0 \\ 0 & \lambda \end{bmatrix}\right) = 0$. |  |
|------|---------|
| Write the characteristic equation with $\lambda$ as the unknown. | |
|------|---------|
| Solve for the two values of $\lambda$. |  |
|------|---------|
| Examine the properties of $\lambda$. Is it real, complex, or imaginary? Are they both positive, negative, or opposite signs? |  |
|------|---------|
| Note that $J_1$ represents the equilibrium $(0,0)$. Use the properties of $\lambda$ to determine the behavior of solutions around $(0,0)$. |  |

\begin{minipage}[t][\textheight-\pagetotal]{0.49\textwidth}
  \begin{itemize}
    \item[b.] Do the same tasks in Part (a) to determine the eigenvalues of $J_2$ and use them to determine the stability of the equilibrium $\left(\frac{5}{4},\frac{15}{7}\right)$.
    \vfill
  \end{itemize}
\end{minipage}
\hfill\vline\hfill
\begin{minipage}[t][\textheight-\pagetotal]{0.49\textwidth}
  \begin{itemize}
    \item[c.] Draw solution curves around each equilibrium on the phase plane and observe their behaviors. Compare your observations to Parts (a) and (b).
```{r echo=FALSE, fig.align='left', message=FALSE, warning=FALSE, out.width="100%", fig.width=3.5,fig.height=3.5}
ode <- function(x,y,parameters){
  with(as.list(c(parameters)),{
    dx <- 3*x-1.4*x*y
    dy <- -y+0.8*x*y
    return(list(dx,dy))
  })
}

PhasePortrait(ode,params=c(),
              xlim=c(-1,2.5),ylim=c(-1,3.5),res=15,
              scale=0.20,rad=0.28,unit=TRUE,
              col="black",xlab="x",ylab="y",title="")  +
  geom_point(data=tibble(x=0,y=0),aes(x=x,y=y),color="red") +
  geom_point(data=tibble(x=1/0.80,y=3/1.40),aes(x=x,y=y),color="red")
```
  \end{itemize}
\end{minipage}
