---
header-includes: |
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhead[LO,L]{\textbf{MTH-321 AB} | Spring 2026}
  \fancyhead[RO,R]{\textbf{Worksheet 25} | Pg. \thepage}
  \fancyfoot[CO,C]{}
  \usepackage{float}
  \usepackage{multirow}
output:
  bookdown::pdf_document2:
    toc: no
    number_section: no
urlcolor: blue
link-citations: yes
bibliography: ../../../references.bib
csl: ../../../mbe.csl
geometry: "left=0.25in,right=0.25in,top=0.75in,bottom=0.75in"
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source(file.path("../../R/functions-mth321.R"))
```

**Name:**

**Collaborators:**

\noindent\rule{\textwidth}{1pt}
**Instructions:** Worksheets are graded mostly on completion and partially on correctness. Please write complete solutions showing explanations and key steps to the following problems, unless it says otherwise.
\noindent\rule{\textwidth}{1pt}

## Find Eigensolutions with Complex Eigenvalues

### 1. Imaginary Eigenvalues

Consider the *linear homogeneous system of 1st-order ODEs with constant coefficients*: $\displaystyle \begin{aligned} x' & = y \\ y' & = -2x \end{aligned}.$

\begin{minipage}[t][\textheight-\pagetotal]{0.49\textwidth}
  \begin{itemize}
    \item[a.] Show that the eigenvalues are purely imaginary $$\lambda_1 = i \sqrt{2} \quad \text{and} \quad \lambda_2 = -i \sqrt{2}.$$
    \vfill
    \item[b.] Show that the corresponding eigenvectors are $$\vec{v}_1 = \begin{bmatrix} -\frac{1}{\sqrt{2}} i \\ 1 \end{bmatrix} \quad \text{and} \quad \vec{v}_2 = \begin{bmatrix} \frac{1}{\sqrt{2}} i \\ 1\end{bmatrix}.$$ Then, write these eigenvectors as a linear combination of its real and imaginary parts.
    \vfill
  \end{itemize}
\end{minipage}
\hfill\vline\hfill
\begin{minipage}[t][\textheight-\pagetotal]{0.49\textwidth}
  \begin{itemize}
    \item[c.] Below is the vector field for the given system. Draw the real and imaginary parts of the eigenvectors on the phase plane.
```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width="100%", fig.width=4,fig.height=4}
ode <- function(x,y,parameters){
  with(as.list(c(parameters)),{
    dx <- y
    dy <- - 2*x
    return(list(dx,dy))
  })
}

PhasePortrait(ode,params=c(),
              xlim=c(-2,2),ylim=c(-2,2),res=13,
              scale=0.20,rad=0.30,unit=TRUE,axis=FALSE,
              col="black",xlab=TeX('$x$'),ylab=TeX('$y$'),title='') + 
  geom_point(data=tibble(x=0,y=0),aes(x=x,y=y),color="red")
```
    \item[d.] Draw a specific solution curve on the direction field with initial condition $(x,y)=(1,0)$. Then, discuss how the parts of the complex eigenvectors help determine the orientation of the ellipse formed by the solution.
    \vfill
  \end{itemize}
\end{minipage}

\newpage

### 2. Complex Eigenvalues

Consider the *linear homogeneous system of 1st-order ODEs with constant coefficients*: $\displaystyle \begin{aligned} x' & = y \\ y' & = -2x-2y \end{aligned}.$

\begin{minipage}[t][\textheight-\pagetotal]{0.49\textwidth}
  \begin{itemize}
    \item[a.] Show that the eigenvalues are complex conjugates $$\lambda_1 = -1 + i \quad \text{and} \quad \lambda_2 = -1 -i.$$
    \vfill
    \item[b.] Show that the corresponding eigenvectors are $$\vec{v}_1 = \begin{bmatrix} -1-i \\ 2 \end{bmatrix} \quad \text{and} \quad \vec{v}_2 = \begin{bmatrix} -1+i \\ 2 \end{bmatrix}.$$ Then, write these eigenvectors as a linear combination of its real and imaginary components.
    \vfill
    \item[c.] Determine the general solution. Here, you can use $\lambda_1$ and $\vec{v}_1$ to form a solution using Euler's formula. Then, use the real and imaginary parts of this solution as two linearly independent solutions to form a general solution.
    \vfill
  \end{itemize}
\end{minipage}
\hfill\vline\hfill
\begin{minipage}[t][\textheight-\pagetotal]{0.49\textwidth}
  \begin{itemize}
    \item[d.] Below is the vector field for the given system. Draw the real and imaginary parts of the eigenvectors on the phase plane.
```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width="100%", fig.width=4,fig.height=4}
ode <- function(x,y,parameters){
  with(as.list(c(parameters)),{
    dx <- y
    dy <- -2*x - 2*y
    return(list(dx,dy))
  })
}

PhasePortrait(ode,params=c(),
              xlim=c(-2,2),ylim=c(-2,2),res=13,
              scale=0.20,rad=0.30,unit=TRUE,axis=FALSE,
              col="black",xlab=TeX('$x$'),ylab=TeX('$y$'),title='') + 
  geom_point(data=tibble(x=0,y=0),aes(x=x,y=y),color="red")
```
  \item[e.] Draw a specific solution curve on the direction field with initial condition $(x,y)=(1,0)$. Then, discuss how the eigenvectors help determine the orientation of the spiral formed by the solution.
  \vfill
  \item[f.] Determine the specific solution given the initial condition $(x,y)=(1,0)$.
  \vfill
  \end{itemize}
\end{minipage}
